---
title: "RNASeq-Differential-Expression of SARS-CoV-2"
author: "Nunez Gonzalez, Laura"
date: "2026-02-11"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    self_contained: true
  pdf_document:
    toc: true
  word_document:
    toc: true
---
<style>
body {
  text-align: justify;
} 

.hanging-indent {
  padding-left: 2em;
  text-indent: -2em;
  margin-bottom: 1.5em; 
}
</style>
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path = "plots/" 
)
```


# Libraries

```{r}
# ==============================================================================
# 1. DATA ACQUISITION AND GENOMIC ANNOTATION
# ==============================================================================

# Genome annotation from Homo sapiens
library(org.Hs.eg.db) 
library(EnsDb.Hsapiens.v86) 

# Infrastructure for representing and manipulating data from GEO and genomic intervals
library(SummarizedExperiment) 
library(GenomicRanges) 

# ==============================================================================
# 2. RNA-seq DATA PREPROCESSING AND READ COUNT NORMALIZATION
# ==============================================================================

# General data manipulation
library(tidyverse)
library(here)
library(dplyr)
library (knitr)
library(kableExtra)
 
#TMM and log2 normalization
library(edgeR)

# ==============================================================================
# 3. EXPLORATORY DATA ANALYSIS
# ==============================================================================

# Visualization 
library(ggplot2)       # Basic visualization tool
library(ggrepel)       # Modification of text over the graphs
library(RColorBrewer)  # Color palettes
library(gridExtra)     # Arrange grid in plots
library(pheatmap)      # Creation of high-quality heat maps
library(dendextend)    # Color palettes in dendograms
library (patchwork)    # Combination of different graphics
library(ComplexUpset)  # Creation of graphics for multiple intersections (UpSet plots)

# ==============================================================================
# 4. DIFFERENTIAL GENE EXPRESSION ANALYSIS: FUNCTIONAL ENRICHMENT & INTERPRETATION
# ==============================================================================

# Statistical frameworks for Differential Expression Genes (DEGs) analysis
library(DESeq2) 
library(EnhancedVolcano) # Creation of high-quality volcano plots of DEGs
library(BiocParallel)
library(clusterProfiler) #Functional enrichment analysis

# Note: Detailed library information is documented in the session_info.txt.
```
# 1. Introduction
This study, published in 2021 following the **SARS-CoV-2 pandemic (McClain et al., 2021)**, explores the activation of biological processes and molecular markers in **peripheral blood samples from infected patients**. These findings are further compared against other respiratory conditions, **including bacterial pneumonia, influenza virus infection, and healthy controls**.

The analytical approach employed is **mRNA transcriptomics using an universal library preparation**. Sequencing was performed on the Illumina® NovaSeq 6000 platform. Stringent quality control filters were applied, including a minimum read length of 100 base pairs and a depth of at least 50 million reads per patient sample.

Collectively, this research provides an exhaustive understanding of the biological—primarily immunological—response associated with each respiratory pathology, enabling a detailed comparative analysis of their shared and distinct molecular signatures. 

# 2. Objectives
1. Data Processing & Replication: Using the raw data deposited in GEO (GSE161731), this project performs a complete end-to-end processing workflow. The analysis is conducted starting from raw counts, replicating the original investigators' rigorous approach to ensure data integrity.
2. Structured Transcriptomic Pipeline: The project follows a systematic and ordered pipeline for analyzing transcriptomic data. This workflow transitions from raw data preprocessing and normalization to a final functional analysis, identifying the dysregulated biological pathways and molecular signatures associated with each respiratory pathology.

# 3. Methodology 
The transcriptomic **workflow was implemented in R (v4.x) using the Bioconductor framework**. Information of packages and versions is available in Session_info.txt file. Raw sequencing counts and clinical metadata were retrieved from the NCBI **Gene Expression Omnibus (GEO) under accession number GSE161731** and integrated into a **SummarizedExperiment object for downstream analysis**. To ensure accurate feature attribution, **genomic coordinates** were mapped to the transcripts, followed by a rigorous data-wrangling process to filter out non-informative variables and standardize gene nomenclature.

To account for **potential library composition biases across the 195 samples, we applied the Trimmed Mean of M-values (TMM) normalization method via the edgeR package**. This "trimming" step was essential to stabilize variance and allow for robust inter-sample comparisons. **Exploratory Data Analysis (EDA) was then performed using Principal Component Analysis (PCA) and unsupervised hierarchical clustering. For the latter, Ward’s Method (Ward.D2)** was utilized to minimize within-cluster variance, providing a high-sensitivity framework for the detection and subsequent removal of technical or **outliers**.

Following quality control, **Differential Expression Analysis (DEA) was conducted using the DESeq2 package**, which employs empirical Bayes shrinkage to estimate dispersion and fold-change. Finally, to characterize the biological significance of the results, we performed Functional Enrichment Analysis. **Overrepresented Gene Ontology (GO)** terms were identified to elucidate the molecular pathways and immunological signatures associated with COVID-19 and bacterial pneumonia relative to healthy controls.

# 4. Results

## 4.1 Data acquisition

Data acquisition is implemented from local archives (downloaded from GEO):
```{r}
data_path <- here("GSE161731")

#Expression matrix
counts <- read.csv(here("GSE161731", "GSE161731_counts.csv.gz"),
                   row.names = 1, check.names = FALSE) #with row.name = 1 we associate all the row data to the gene (which is detailed in the first column)

#Metadata
metadata <- read.csv(here("GSE161731", "GSE161731_key.csv.gz"),
                     header = TRUE, row.names = 1, check.names = FALSE) #same argument, but in this case we are associating sample id (patient) with the remaining data

#Specific selection of the intersection about colnames from counts and rownames from metadata
common_samples <- intersect(colnames(counts), rownames(metadata))
counts_clean <- as.matrix(counts[, common_samples]) #sorted based on common_samples
metadata_clean <- metadata[common_samples, ]
```

Addition of chromosomic location: 
```{r}
# Removal of different gene versions
rownames(counts_clean) <- gsub("\\..*$", "", rownames(counts_clean))

#Annotation calling
edb <- EnsDb.Hsapiens.v86
genes_coords <- genes(edb)

common_genes <- intersect(rownames(counts_clean), names(genes_coords))

counts_final <- counts_clean[common_genes, ]
coords_final <- genes_coords[common_genes]
metadata_final <- metadata_clean  
```

Now, we can create the object: **summarizedExperiment**
```{r}
se <- SummarizedExperiment(
  assays = list(counts = counts_final),
  colData = DataFrame(metadata_final),
  rowRanges = coords_final
)

#We include metadata information (empty by defect)
metadata(se) <- list(
  GEO_accession = "GSE161731",
  organism = "Homo sapiens",
  assay_type = "RNA-seq"
)

#Definitive object: summarizedExperiment
se
```

## 4.2 RNA-seq Data Preprocessing and Read Count Normalization

### 4.2.1 Data preprocessing 
```{r}
# DUPLICATES EXTRACTION AND REMOVAL
# We will maintain only the first entry of each duplicated sample
metadata_raw <- as.data.frame(colData(se)) #We convert the colData in a normal dataframe for easy manipulation 
metadata_unique <- metadata_raw[!duplicated(rownames(metadata_raw)), ]
rownames(metadata_unique) <- rownames(metadata_raw)[!duplicated(rownames(metadata_raw))]

# Selection of specific cohorts
target_cohorts <- c("COVID-19", "Bacterial", "healthy")
metadata_cohorts <- metadata_unique[metadata_unique$cohort %in% target_cohorts, ]

# Standarization of special characters
colnames(metadata_cohorts) <- gsub("[[:space:]\\-\\\\/]", "_", colnames(metadata_cohorts)) #Standarized colnames, withouth hyphens, spaces and slashes (replaced by underscore)

#Now, we will clean the data inside the columns (in the character columns): 
metadata_cohorts <- metadata_cohorts %>%
  mutate(across(where(is.character), ~ gsub("\\s|\\-|/", "_", .x)))

#We eliminate all the characters in age column
metadata_cohorts$age <- as.numeric(gsub("[^0-9]", "", metadata_cohorts$age))

#Get the summarizedExperiment with all these changes included
keep_samples <- rownames(metadata_cohorts)
se_clean <- se[, keep_samples]
colData(se_clean) <- DataFrame(metadata_cohorts)

cohort_counts <- table(colData(se_clean)$cohort)
tabla_df <- as.data.frame(cohort_counts)
colnames(tabla_df) <- c("Clinical Cohort", "Number of Samples")
kable(tabla_df, caption = "Table 1: Sample distribution across clinical cohorts.")
```
Now, we have 3 clean cohort which contains 19 healthy individuals, 23 who suffered from bacterial pneumonia and 77 patients from COVID-19. 

Once the data has been correctly preprocessed, we must **remove all the genes with low expression (probable biological noise)**. 

```{r}
# Removal of genes with <10 total counts 
keep <- rowSums(assay(se_clean, "counts")) >= 10

#Removal of these genes in which the 90% of the values correspond to zero
keep <- keep & rowSums(assay(se_clean, "counts") == 0) <= 0.9 * ncol(se_clean)

se_filtered <- se_clean[keep, ] #We maintain the aforementioned files (TRUE under keep variable and all the columns)
print(paste("Genes after data cleaning processing:", sum(keep), "Final genes for the study:", nrow(se_filtered)))
```
At the beginning we had near 60K genes. Now, we will work only with near to 30K (**50% of the genes are noise to the study**). 

### 4.2.2 TMM and log Normalization
Following author's criteria, we will employ TMM (Trimmed Mean of M-values) normalization (Robinson & Oshlack, 2010) . With this methodology, the most central values will be averaging and the extreme values will be dropped out. Normalized counts were then transformed to log~2~CPM with pseudo-count (prior.count=1) with this criteria: 

**TMM normalization**

1. **For each sample vs. reference** (calculated by edgeR):
   - Compute $log_2$(count$_{\text{sample}}$/count$_{\text{reference}}$) **per gene**
   - **Trim** extreme values (75% outliers removed)  
   - Average the central 25% → **TMM factor**

2. **Normalize**: $counts_{\text{TMM}} = raw\_counts / TMM\_factor$

3. **log2CPM**: $logCPM = log_2\left(\frac{counts_{\text{TMM}} \times 10^6}{library\_size} + 1\right)$

This double-normalization corrects both **composition bias** (TMM) and **library size** (CPM), stabilizing variance for downstream PCA/clustering.

TMM normalization: 
   For each sample vs. reference sample (edgeR)
   - TMM factor considerate the number of reads of each gene
   - Calculate $log2(raw counts/TMM factor])$ for each single sample
   - Remove extreme values: atypical data
   - Average 25% of central values

$normalized_counts = raw counts / TMM factor$

$logCPM = log2( (normalized_counts * 1e6) / 1e6 + prior.count )$

```{r}
# 1. Extraction of raw data (counts) to create a new object: DEGList
counts_matrix <- assay(se_filtered, "counts")
dge <- DGEList(counts = counts_matrix)
dge <- calcNormFactors(dge, method = "TMM") #TMM is already included in this package and ready to use

# Logarithmic transformation
logCPM <- cpm(dge, log = TRUE, prior.count = 1)

# We save this information as new assay
assay(se_filtered, "TMM_log2") <- logCPM
```

## 4.3 Exploratory data analysis (EDA)

### 4.3.1 Visualization tools
#### 4.3.1.1 PCA (Principal Component Analysis)
```{r 1.PCA plot}
# Data extraction from TMM_log2
plot_data <- assay(se_filtered, "TMM_log2")
pca_res <- prcomp(t(plot_data), scale. = TRUE)

# Data frame for ggplot
pca_df <- data.frame(
  PC1 = pca_res$x[,1],
  PC2 = pca_res$x[,2],
  cohort = colData(se_filtered)$cohort,
  batch = colData(se_filtered)$batch,
  gender = colData(se_filtered)$gender,
  sample = colnames(se_filtered)
)

# PCA
ggplot(pca_df, aes(PC1, PC2, color = cohort)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "PCA - TMM_log2 normalized", 
       x = paste0("PC1 (", round(summary(pca_res)$importance[2,1]*100,1), "%)"),
       y = paste0("PC2 (", round(summary(pca_res)$importance[2,2]*100,1), "%)")) +
  theme_minimal()
```

Visual analysis of the gene expression profiles reveals a **clear segregation of the bacterial infection samples**. In contrast, considerable **overlap is observed between the transcripts of COVID-19 patients and healthy controls**.

#### 4.3.1.2 MDS (Multidimensional Scaling)
```{r 2.MDS plot}
# MDS 
palette <- c("Bacterial" = "red", "COVID_19" = "green", "healthy" = "blue")
cohort_factor <- factor(colData(se_filtered)$cohort)
palette_mds <- palette[cohort_factor]

plotMDS(dge, 
        col = palette_mds,                    
        labels = paste(colData(se_filtered)$cohort, 
                      colData(se_filtered)$batch, sep="_"),
        main = "MDS: RED=Bacterial, GREEN=COVID, BLUE=Healthy")
```

Both analyses (PCA and MDS) demonstrate a **consistent sample segregation**, confirming the robustness of the overall clustering pattern. 

#### 4.3.1.3 Heatmap

```{r 3.Heatmap, fig.width=12, fig.height=4}
tmm_data <- assay(se_filtered, "TMM_log2")
tmm_clean <- tmm_data; tmm_clean[is.infinite(tmm_data) | is.na(tmm_data)] <- 0

# Top 1000 genes variables
rv <- apply(tmm_clean, 1, var, na.rm = TRUE); rv[is.na(rv)] <- 0
top_genes <- order(rv, decreasing = TRUE)[1:1000]

annotation_col <- data.frame(Cohort = colData(se_filtered)$cohort)
rownames(annotation_col) <- colnames(tmm_clean)
annotation_colors <- list(Cohort = c("Bacterial" = "red", "COVID_19" = "green", "healthy" = "blue"))
pheatmap(tmm_clean[top_genes, ], 
         annotation_col = annotation_col,
         annotation_colors = annotation_colors,
         show_rownames = FALSE,
         show_colnames = FALSE,
         main = "RED=Bacterial, GREEN=COVID, BLUE=Healthy")
```

The hierarchical clustering dendrogram reveals **distinct separation of the Bacterial cohort** (red annotation bar), while **COVID-19 and healthy samples show partial overlap**. 
This confirms previous visualization tools, such as PCA.

### 4.3.2 Atypical data identification

We select **Ward method to calculate minimum variance inside each cluster** (version D2).
```{r 4.Outlier visualization}
# Distance calculation
dist_matrix <- dist(t(tmm_clean))
hclust_res <- hclust(dist_matrix, method = "ward.D2")

# Plot to visualize outliers
plot(hclust_res, labels = FALSE, main = "Clustering - Possible Outliers")
abline(h = 30, col = "red", lwd = 2)  # Cutoff
rect.hclust(hclust_res, k = 5, border = "red")
```

With this method, 5 different clusters were created. Now, we must see the possible correspondence of these clusters with biological cohorts. 
```{r}
# We define the previous clusters
clusters_5 <- cutree(hclust_res, k = 5)

# We extract cohort data from summarizedExperiment
comparison_df <- data.frame(
  Sample = names(clusters_5),
  Cluster = as.factor(clusters_5),
  Cohort = colData(se_filtered)$cohort
)

# Table
contingency_table <- table(Cluster = comparison_df$Cluster, Cohort = comparison_df$Cohort)
```

```{r 5.Cluster distribution by cohort}
ggplot(comparison_df, aes(x = Cluster, fill = Cohort)) +
  geom_bar(position = "fill") + # "fill" muestra proporciones (0 a 1)
  theme_minimal() +
  scale_fill_manual(values = c("Bacterial" = "#E41A1C", "COVID_19" = "#4DAF4A", "healthy" = "#377EB8")) +
  labs(title = "Cluster distribution by cohort",
       subtitle = "Purity evaluation of the 5 clusters",
       x = "Cluster(n=5)",
       y = "Sample proportion") +
  theme(legend.position = "bottom")
```

This graph shows a clear overlap in COVID-19 and control patients in clusters from 1 to 3. However, bacterial pneumonia patients are clearly separated in clusters 4 and 5.

```{r}

# Identification of small clusters (n <= 2)
counts_por_cluster <- table(clusters_5)
clusters_outliers <- names(counts_por_cluster[counts_por_cluster <= 2])

# Extraction of the samples within this criteria
final_outliers <- names(clusters_5[clusters_5 %in% clusters_outliers])

# Cleaning up
se_clean <- se_filtered[, !(colnames(se_filtered) %in% final_outliers)]

# Verification
message("Original samples: ", ncol(se_filtered))
message("Removed samples: ", length(final_outliers))
message("Final samples: ", ncol(se_clean))
```
After this short verification, we will continue the identification of atypical data. Based on Spearman correlation matrix for each individual. The patients with -2SD or +2SD with respect to the mean will be considered as outlier: 
```{r}
sample_cor <- cor(tmm_clean, method = "spearman")
mean_cor <- colMeans(sample_cor)

# Threshold 
threshold_cor <- mean(mean_cor) - 2 * sd(mean_cor)
outliers_num <- names(mean_cor[mean_cor < threshold_cor])

cat("Outliers identified:", length(outliers_num), "These samples will be removed for subsequent posterior analysis.")

#Outliers removal
se_clean <- se_filtered[, !(colnames(se_filtered) %in% outliers_num)]
message(paste("Analysis contains", ncol(se_clean), "samples."))
```

### 4.3.3 Confounding variables

We will visualize PCA by different confounding variables
```{r}
# Variables to consider
pca_df$batch        <- colData(se_filtered)$batch
pca_df$age <- as.numeric(colData(se_filtered)$age)
pca_df$race <- colData(se_filtered)$race
pca_df$gender <- colData(se_filtered)$gender
pca_df$hospitalized <- colData(se_filtered)$hospitalized
pca_df$onset        <- colData(se_filtered)$time_since_onset 

# All the variables employed must be factors
# Healthy controls must be the reference cohort
se_filtered$cohort <- factor(se_filtered$cohort)
se_filtered$cohort <- relevel(se_filtered$cohort, ref = "healthy")

se_filtered$batch  <- factor(se_filtered$batch)
se_filtered$race   <- factor(se_filtered$race)
se_filtered$age <- scale(as.numeric(as.character(se_filtered$age))) 

#We have 42 patients with no data for variables: hospitalized and onset. Probably, they are healthy controls
contingency_table_hosp <- table(pca_df$cohort, is.na(pca_df$hospitalized))
table_mat <- as.matrix(contingency_table_hosp) #Matrix conversion to manage the internal names
colnames(table_mat) <- c("Complete Data", "Missing Data (NA)")
kable(table_mat, #table generation
      caption = "Table 2: Analysis of missing clinical data (Hospitalized/Onset) across cohorts.",
      booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = FALSE, 
                position = "center")
```
All patients were hospitalized, whereas none of the healthy controls were. Consequently, **hospitalization status is a nested variable and should not be included as a confounding factor** in the analysis of the full cohort.

```{r 6.PCAs by confounding variables, fig.width=10, fig.height=10}

#PCA by batch
p_batch <- ggplot(pca_df, aes(PC1, PC2, color = batch)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "PCA by Batch", x = "PC1", y = "PC2") +
  theme_minimal()

# PCA by race
p_race <- ggplot(pca_df, aes(PC1, PC2, color = race)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "PCA by Race", x = "PC1", y = "PC2") +
  theme_minimal()

# PCA by sex
p_sex <- ggplot(pca_df, aes(PC1, PC2, color = gender)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(title = "PCA by sex", x = "PC1", y = "PC2") +
  theme_minimal()

# PCA by age (continuous variable)
p_age <- ggplot(pca_df, aes(PC1, PC2, color = age)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_viridis_c() +
  labs(title = "PCA by Age", x = "PC1", y = "PC2") +
  theme_minimal()

# Plots
pca_final_grid <- (p_batch | p_race) / 
                  (p_age | p_sex) +
  plot_annotation(
    title = 'Exploratory Data Analysis: PCA Multivariable assessment',
    subtitle = 'Analysis of potential biological and technical confounding factors',
    caption = 'Dataset: RNA-Seq COVID-19 Study | 195 Samples'
  ) & 
  theme(plot.title = element_text(size = 18, face = "bold"))
pca_final_grid
```

Furthermore, visual inspection reveals that Batch 2 is predominantly localized to the right side of the plot, while Batch 1 exhibits a broader distribution. **Including batch as a covariate in the model is highly recommended**, as it is a standard approach to mitigate potential technical artifacts arising from the sequencing workflow Hicks et al. (2018). 

Regarding **race**, White and African American individuals constitute the vast majority of the cohort. There appears to be a **clear distinction between these two racial groups**.

There is **no clear separation between males and females**; thus, biological sex was not considered a confounding variable. In contrast, noticeable differences are observed regarding the age of the study subjects. Consequently, **age was included as a covariate in the model**.

## 4.4 Differential Gene Expression (DGE) Analysis: Functional Enrichment and Interpretation

We will implement a **generalized linear model (GLM)** incorporating race, age, and batch as confounding covariates to isolate their effects from the cohort variable. Our selected method is **DESeq2**.

The primary objective of creating the DESeqDataSet object is to **evaluate differential gene expression across cohorts while ensuring the mitigation of technical and biological noise** caused by these confounding variables.

```{r dsa, cache = TRUE, cache.path = "c/"}
register(SnowParam(workers = 6))
# Matrix construction
dds <- DESeqDataSet(se_filtered, design = ~ batch + race + age + cohort)

# Execution of the analysis (dispersion estimation and model fitting
# Shrinkage is included in this step
dds <- DESeq(dds, parallel= T)
```

Next, we specify the contrast vectors to compare COVID-19 vs. healthy and Bacterial pneumonia vs. healthy, allowing us to extract the differential expression results for each condition.
```{r}
# Contrast: COVID-19 vs Healthy
res_covid <- results(dds, 
                     contrast = c("cohort", "COVID_19", "healthy"),
                     alpha = 0.05) # Umbral de significación

# Contrast: Bacterial vs Healthy
res_bacterial <- results(dds, 
                         contrast = c("cohort", "Bacterial", "healthy"),
                         alpha = 0.05)

# Summaries
message("DEGs: COVID-19 vs Healthy")
message("DEGs: Bacterial vs Healthy")
```

Differential gene expression analysis using the **negative binomial distribution (DESeq2) with Benjamini-Hochberg correction by FDR** revealed significant transcriptomic alterations in both infection cohorts compared to the healthy control group.

The **bacterial cohort exhibited a more extensive transcriptional response, with approximately 26% of genes showing differential expression ($padj < 0.05$)**, compared to **6.8% in the COVID-19 group**. Notably, the COVID-19 analysis required a more stringent independent filtering, with 47% of genes excluded due to low mean counts. 

**Outlier detection remained consistent across both comparisons, accounting for approximately 2%** of the total features. These results highlight a distinct molecular signature for each pathogen, with bacterial pneumonia triggering a broader systemic gene dysregulation than COVID-19 in this dataset.

### 4.4.1 Visualization of DEG

```{r 7.MAplot}
plotMA_covid19 <- plotMA(res_covid, ylim=c(-5, 5), 
       main="Differential Expression distribution: COVID-19 vs. healthy")
# Reference in log2FoldChange = 1
abline(h=c(-1, 1), col="green4", lty=5) 

plotMA_bact<- plotMA(res_bacterial, ylim=c(-5, 5), 
       main="Differential Expression distribution: bacterial pneumonia vs. healthy")
# Reference in log2FoldChange = 1
abline(h=c(-1, 1), col= "red2", lty=5) 
```

The **comparative differential expression analysis reveals distinct transcriptional landscapes between bacterial and viral etiologies**. While bacterial pneumonia induces a massive and symmetric inflammatory response characterized by a high number of significantly dysregulated genes, the COVID-19 signature appears more targeted. Notably, the COVID-19 profile exhibits a prominent cluster of down-regulated genes among high-abundance transcripts, suggesting a specific suppressive effect on host cellular processes.

```{r 8.Volcano plot, fig.width=12, fig.height=8}
# Volcano plot for COVID-19 vs Healthy
volcano_covid19 <- EnhancedVolcano(res_covid,
                lab = rownames(res_covid),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'COVID-19 vs Healthy',
                pCutoff = 0.05,
                FCcutoff = 1.5,
                pointSize = 2.0,
                labSize = 4.0,
                col=c('grey', 'grey', 'yellow2', 'blue'),
                legendLabels=c('NS', 'Log2 FC', 'Adjusted p-value', 'Both'),
                caption = 'FC cutoff: 1.5; p-value cutoff: 0.05')

# Volcano plot for Bacterial vs Healthy
volcano_bacterial <- EnhancedVolcano(res_bacterial,
                lab = rownames(res_bacterial),
                x = 'log2FoldChange',
                y = 'padj',
                title = 'Bacterial Pneumonia vs Healthy',
                pCutoff = 0.05,
                FCcutoff = 1.5,
                pointSize = 2.0,
                labSize = 4.0,
                col=c('grey', 'grey', 'yellow2', 'blue'),
                legendLabels=c('NS', 'Log2 FC', 'Adjusted p-value', 'Both'),
                caption = 'FC cutoff: 1.5; p-value cutoff: 0.05')
volcano_covid19 | volcano_bacterial
```

To gain a overview of coincidences between the two cohorts we generate an upset plot. We use a threshold of foldchange 1.5 and alpha = 0.05 to represente Differential Expressed Genes. 

```{r 9.Upset plot}
# Threshold definition
lfc_threshold <- log2(1.5)

# Gene list
list_input <- list(
  `COVID UP`   = rownames(res_covid[which(res_covid$padj < 0.05 & res_covid$log2FoldChange > lfc_threshold), ]),
  `COVID DOWN` = rownames(res_covid[which(res_covid$padj < 0.05 & res_covid$log2FoldChange < -lfc_threshold), ]),
  `Bact UP`    = rownames(res_bacterial[which(res_bacterial$padj < 0.05 & res_bacterial$log2FoldChange > lfc_threshold), ]),
  `Bact DOWN`  = rownames(res_bacterial[which(res_bacterial$padj < 0.05 & res_bacterial$log2FoldChange < -lfc_threshold), ])
)

# Binary vector 
all_genes <- unique(unlist(list_input))
upset_df <- data.frame(gene = all_genes)
for(set_name in names(list_input)) {
  upset_df[[set_name]] <- as.integer(upset_df$gene %in% list_input[[set_name]])
}

# colNames define the genes
sets_indices <- names(list_input)

#Upset plot
upset(
  upset_df,
  sets_indices, # Vector with colNames
  name = "Groups",
  width_ratio = 0.1,
  queries = list(
    upset_query(set = "COVID UP", fill = "#E41A1C"),
    upset_query(set = "Bact UP", fill = "#A60606"),
    upset_query(set = "COVID DOWN", fill = "#377EB8"),
    upset_query(set = "Bact DOWN", fill = "#0D4673")
  ),
  base_annotations = list(
    'Intersection size' = intersection_size(counts = TRUE)
  ),
  set_sizes = upset_set_size() + theme(axis.text.x = element_text(angle = 90))
) + labs(
    title = "Stratified Expression Overlap (FC > 1.5)",
    subtitle = "Comparing Directional Response: COVID-19 vs Bacterial Pneumonia"
)
```

The **intersection analysis confirms a more extensive transcriptional landscape in Bacterial pneumonia compared to COVID-19** (both in up-regulated and down-regulated genes) relative to healthy controls. Furthermore, the intersections reveal that **overlaps between both datasets occur primarily in a concordant manner**: genes up-regulated in COVID-19 tend to be up-regulated in the bacterial cohort, with a similar pattern observed for down-regulated genes.

These results identify a common molecular response to respiratory infections, highlighting **potential biomarkers for general lung inflammation and host defense mechanism**.

## 4.5 Functional enrichment analysis

We select the upregulated genes from COVID-19 and pneumonia patients (over healthy controls) to study the main mechanisms and physiological pathways activated under these conditions. 

```{r 10.Functional enrichment by biological process, fig.width=12, fig.height=8}
covid_up <- list_input[["COVID UP"]]
bact_up  <- list_input[["Bact UP"]]

covid_up_clean <- as.character(covid_up)
bact_up_clean  <- as.character(bact_up)

enr_covid <- enrichGO(gene          = covid_up_clean,
                      OrgDb         = org.Hs.eg.db,
                      keyType       = 'ENSEMBL', 
                      ont           = "BP",     
                      pAdjustMethod = "BH",
                      pvalueCutoff  = 0.05, 
                      qvalueCutoff  = 0.2)  

enr_bact <- enrichGO(gene          = bact_up_clean,
                     OrgDb         = org.Hs.eg.db,
                     keyType       = 'ENSEMBL',
                     ont           = "BP",
                     pAdjustMethod = "BH",
                     pvalueCutoff  = 0.05,
                     qvalueCutoff  = 0.2)

#Visualization of the most enriched biological process in dotplot
enr_covid <- setReadable(enr_covid, OrgDb = org.Hs.eg.db, keyType="ENSEMBL") #Transformation of gene nomenclature
enr_bact  <- setReadable(enr_bact, OrgDb = org.Hs.eg.db, keyType="ENSEMBL")

# Plots
dot_covid <- dotplot(enr_covid, showCategory=15) + 
  ggtitle("Enriched Biological Processes: COVID-19") +
  theme_minimal()

dot_bact <- dotplot(enr_bact, showCategory=15) + 
  ggtitle("Enriched Biological Processes: Bacterial") +
  theme_minimal()

dot_covid | dot_bact
```

**Functional enrichment analysis reveals deep contrast between the two infectious etiologies despite their shared clinical presentation**. In COVID-19, the host response is predominantly characterized by immediate immune activation, specifically through immunoglobulin-mediated responses and B-cell-mediated cellular immunity. To a lesser extent, pathways related to cell division are also enriched, potentially reflecting lymphocyte proliferation.

In contrast, the bacterial infection signature is directly linked to **cellular structural and degradative processes, specifically autophagy, vesicle-mediated transport, and endocytosis**. These pathways likely represent the host's attempt to internalize and degrade bacteria.

# 5. Discussion
The role of **Bioconductor in gene expression analysis** is invaluable for modern biomedical research. Currently, scientific standards require researchers to **deposit high-throughput sequencing data in public repositories such as the GEO (Gene Expression Omnibus)** Database. These datasets are typically available in both raw and processed formats, providing a transparent foundation for the scientific community.

For any bioinformatician the ability to interpret and re-analyze these studies is an essential skill. Public data serves as a powerful tool for comparative and complementary analysis, allowing for the validation of in-house findings against established benchmarks. Furthermore, leveraging the current state-of-the-art through existing experiments significantly optimizes research resources, reducing the need for redundant laboratory testing and accelerating the discovery of disease-specific molecular signatures.

In this study, among the top 15 enriched biological processes in COVID-19 and bacterial pneumonia, there is **minimal overlap between the two conditions**. This suggests that the molecular mechanisms driving the host's defense are highly pathogen-specific, offering a clear transcriptomic basis for differential diagnosis.

# 6. Reproducibility statement
All raw transcriptomic data and corresponding clinical metadata analyzed in this study are publicly available through the **NCBI Gene Expression Omnibus (GEO) under the accession number GSE161731**. This ensures that the foundational evidence for our findings remains accessible for independent verification.

The complete analytical pipeline, including the **R Markdown (.Rmd) source code, pre-processing scripts, and high-resolution figures, is hosted on GitHub (https://github.com/lnungon/RNASeq-Differential-Expression-SARS-CoV-2)**. The repository is structured to allow for full computational replication of the results, from raw count processing to functional enrichment.

To ensure computational reproducibility, detailed information regarding the R environment, including package versions and system dependencies, is documented in the accompanying session_info.txt file.
```{r}
print(sessionInfo(), locale = FALSE)
```


# 7. Bibliography
::: {.hanging-indent}
Hicks, S. C., Townes, F. W., Teng, M., & Irizarry, R. A. (2018). Missing data and technical variability in single-cell RNA-sequencing experiments. Biostatistics, 19(4), 562–578. https://doi.org/10.1093/biostatistics/kxx053 
:::

::: {.hanging-indent}
McClain, M. T., Constantine, F. J., Henao, R., Liu, Y., Tsalik, E. L., Burke, T. W., Steinbrink, J. M., Petzold, E., Nicholson, B. P., Rolfe, R., Kraft, B. D., Kelly, M. S., Saban, D. R., Yu, C., Shen, X., Ko, E. M., Sempowski, G. D., Denny, T. N., Ginsburg, G. S., … Woods, C. W. (2021). Dysregulated transcriptional responses to SARS-CoV-2 in the periphery. Nature Communications, 12(1), 1079. https://doi.org/10.1038/s41467-021-21289-y
:::

::: {.hanging-indent}
McClain, M. T., Constantine, F. J., Henao, R., Liu, Y., Tsalik, E. L., Burke, T. W., Steinbrink, J. M., Petzold, E., Nicholson, B. P., Rolfe, R., Kraft, B. D., Kelly, M. S., Saban, D. R., Yu, C., Shen, X., Ko, E. M., Sempowski, G. D., Denny, T. N., Ginsburg, G. S., … Woods, C. W. (2021). Dysregulated transcriptional responses to SARS-CoV-2 in the periphery. Nature Communications, 12(1), 1079. https://doi.org/10.1038/s41467-021-21289-y
:::

